---
title: "Dimensionality Reduction"
author: "James Spurrier"
date: "`r Sys.Date()`"
output: html_document
editor:
  mode: source
---

```{r}
library(tidyverse)
library(DESeq2)
library(ComplexHeatmap)
library(PCAtools)
library(umap)
library(RColorBrewer)
library(circlize)
```

## Import raw counts files

```{r}
# SRA study IDs
studyID_1 <- paste0("SRP104161")
studyID_2 <- paste0("SRP241904")
studyID_3 <- paste0("SRP406132")
studyID_4 <- paste0("SRP482110")

# Path to metadata with sample information
metadataFileName <- paste0("metadata_pcaUmap.csv")
metadataDir <- file.path("metadata/exdata")
#metadataDir <- file.path("~/ThesisProject/metadata/exdata")
metadata <- as.data.frame(read_csv(file = file.path(metadataDir, metadataFileName)))

# File names for counts files
countsFileName_1 <- paste0("featurecounts_", studyID_1, ".RData")
countsFileName_2 <- paste0("featurecounts_", studyID_2, ".RData")
countsFileName_3 <- paste0("featurecounts_", studyID_3, ".RData")
countsFileName_4 <- paste0("featurecounts_", studyID_4, ".RData")

# Directory to counts files
countsDir <- file.path("/media/james/META70006/ThesisProject/data/exdata")
#countsDir <- file.path("~/data/exdata")

# This function loads Rdata objects that have identical environment names (i.e "featurecounts")
loadRData <- function(fileName){
#loads an RData file, and returns it
    load(fileName)
    get(ls()[ls() != "fileName"])
}

# Load rawcounts
featurecounts_1 <- loadRData(file = file.path(countsDir, studyID_1, countsFileName_1))
featurecounts_2 <- loadRData(file = file.path(countsDir, studyID_2, countsFileName_2))
featurecounts_3 <- loadRData(file = file.path(countsDir, studyID_3, countsFileName_3))
featurecounts_4 <- loadRData(file = file.path(countsDir, studyID_4, countsFileName_4))
```

```{r}
# Set and create output directories
resultsDir <- file.path("results/exdata/pcaUmap")
#resultsDir <- file.path("~/ThesisProject/results/exdata/pcaUmap")
dir.create(resultsDir)
```

## Combine individual count matrices to single file

```{r}
# Remove all suffix from gene id and column names
counts1 <- as.data.frame(featurecounts_1$counts)
rownames(counts1) <- gsub("[.].*","",rownames(counts1))
colnames(counts1) <- gsub("[.].*","",colnames(counts1))
counts1 <- rownames_to_column(counts1, var = "gene_id")

counts2 <- as.data.frame(featurecounts_2$counts)
rownames(counts2) <- gsub("[.].*","",rownames(counts2))
colnames(counts2) <- gsub("[.].*","",colnames(counts2))
counts2 <- rownames_to_column(counts2, var = "gene_id")

counts3 <- as.data.frame(featurecounts_3$counts)
rownames(counts3) <- gsub("[.].*","",rownames(counts3))
colnames(counts3) <- gsub("[.].*","",colnames(counts3))
counts3 <- rownames_to_column(counts3, var = "gene_id")

counts4 <- as.data.frame(featurecounts_4$counts)
rownames(counts4) <- gsub("[.].*","",rownames(counts4))
colnames(counts4) <- gsub("[.].*","",colnames(counts4))
counts4 <- rownames_to_column(counts4, var = "gene_id")

# List all counts matrices
counts <- list(SRP104161=counts1,
               SRP241904=counts2,
               SRP406132=counts3,
               SRP482110=counts4)
```

```{r}
# Reduce list to single matrix
merged_counts <- Reduce(function(x, y) merge(x, y, by = "gene_id", all = TRUE), counts)
merged_matrix <- as.matrix(column_to_rownames(merged_counts, var = "gene_id"))
```

```{r}
# Match rownames in metadata with colnames in counts matrix
metadata <- column_to_rownames(metadata, var = "Run")
genomic_idx <- match(rownames(metadata), colnames(merged_matrix))
matrix_ordered  <- merged_matrix[ , genomic_idx]
all(rownames(metadata) == colnames(matrix_ordered))
```

```{r}
# Filter counts by cpm
cpms <- edgeR::cpm(matrix_ordered)
keep <- rowSums(cpms >= 1 ) >= 2
cpms <- cpms[keep,]
```

## Principal Component Analysis of cpms

```{r}
p <- pca(cpms, metadata = metadata, removeVar = 0.1)
```

```{r}
# Screeplot
scree.pdf <- paste0("Scree_cpm1_filtered.pdf")
pdf(file = file.path(resultsDir, scree.pdf), width = 12, height = 8)
print(screeplot(p, axisLabSize = 18, titleLabSize = 22, components = 1:10))
dev.off()

scree.svg <- paste0("Scree_cpm1_filtered.svg")
pdf(file = file.path(resultsDir, scree.svg), width = 12, height = 8)
print(screeplot(p, axisLabSize = 18, titleLabSize = 22, components = 1:10))
dev.off()
```

```{r}
# PCs to plot
x <- paste0("PC1")
y <- paste0("PC2")
# PCA plot
pcabiplot.pdf <- paste0("PCA_biplot_cpm1_filtered.pdf")
pdf(file = file.path(resultsDir, pcabiplot.pdf), width = 12, height = 12)
print(biplot(p, x = x, y = y, colby = "Source", shape = "Study", hline = 0, vline = 0,
             legendPosition = 'right', 
             xlab = paste0(x, ", ", round(p$variance[x], digits = 2), "%"), 
             ylab = paste0(y, ", ", round(p$variance[y], digits = 2), "%"), lab = NULL))
dev.off()

pcabiplot.svg <- paste0("PCA_biplot_cpm1_filtered.svg")
svg(file = file.path(resultsDir, pcabiplot.svg), width = 12, height = 12)
print(biplot(p, x = x, y = y, colby = "Source", shape = "Study", hline = 0, vline = 0,
             legendPosition = 'right', 
             xlab = paste0(x, ", ", round(p$variance[x], digits = 2), "%"), 
             ylab = paste0(y, ", ", round(p$variance[y], digits = 2), "%"), lab = NULL))
dev.off()
```

```{r}
pairsplot.pdf <- paste0("pairs_plot_PCA1-5_cpm1_filtered.pdf")
pdf(file = file.path(resultsDir, pairsplot.pdf), width = 12, height = 12)
print(pairsplot(p, colby = "Source", shape = "Study", components = 1:5))
dev.off()

pairsplot.svg <- paste0("pairs_plot_PCA1-5_cpm1_filtered.svg")
pdf(file = file.path(resultsDir, pairsplot.svg), width = 12, height = 12)
print(pairsplot(p, colby = "Source", shape = "Study", components = 1:5))
dev.off()
```

```{r}
rm(p)
```

## Principle component analysis of variance stabilising transformed data

```{r}
# Create deseq dataset
coldata <- metadata
dds <- DESeqDataSetFromMatrix(countData = matrix_ordered,
                              colData = coldata,
                              design = ~ Metastatic)
# Set reference for comparison
dds$Metastatic <- relevel(dds$Metastatic, ref = "No")

# Prefilter low counts
dds <- estimateSizeFactors(dds)
keep <- rowSums( counts(dds, normalized=TRUE) >= 1 ) >= 2
dds <- dds[keep,]

# Run Deseq
dds <- DESeq(dds)

# Extract variance stabilising transformed count from assay field of dds object
vsd <- assay(vst(dds))
```

```{r}
p <- pca(vsd, metadata = colData(dds), removeVar = 0.1)
```

```{r}
# Screeplot
scree.pdf <- paste0("Scree_vst.pdf")
pdf(file = file.path(resultsDir, scree.pdf), width = 12, height = 8)
print(screeplot(p, axisLabSize = 18, titleLabSize = 22, components = 1:10))
dev.off()

scree.svg <- paste0("Scree_vst.svg")
pdf(file = file.path(resultsDir, scree.svg), width = 12, height = 8)
print(screeplot(p, axisLabSize = 18, titleLabSize = 22, components = 1:10))
dev.off()
```

```{r}
# PCA plot
x <- paste0("PC1")
y <- paste0("PC2")

pcabiplot.pdf <- paste0("PCA_biplot_vst.pdf")
pdf(file = file.path(resultsDir, pcabiplot.pdf), width = 10, height = 6)
print(biplot(p, x = x, y = y, colby = "Source", shape = "Study", hline = 0, vline = 0,
             legendPosition = 'right', 
             xlab = paste0(x, ", ", round(p$variance[x], digits = 2), "%"), 
             ylab = paste0(y, ", ", round(p$variance[y], digits = 2), "%"), lab = NULL))
dev.off()

pcabiplot.svg <- paste0("PCA_biplot_vst.svg")
svg(file = file.path(resultsDir, pcabiplot.svg), width = 10, height = 6)
print(biplot(p, x = x, y = y, colby = "Source", shape = "Study", hline = 0, vline = 0,
             legendPosition = 'right', 
             xlab = paste0(x, ", ", round(p$variance[x], digits = 2), "%"), 
             ylab = paste0(y, ", ", round(p$variance[y], digits = 2), "%"), lab = NULL))
dev.off()
```

```{r}
pairsplot.pdf <- paste0("pairs_plot_PCA1-5_vst.pdf")
pdf(file = file.path(resultsDir, pairsplot.pdf), width = 10, height = 10)
print(pairsplot(p, colby = "Source", shape = "Study", components = 1:5))
dev.off()

pairsplot.svg <- paste0("pairs_plot_PCA1-5_vst.svg")
pdf(file = file.path(resultsDir, pairsplot.svg), width = 10, height = 10)
print(pairsplot(p, colby = "Source", shape = "Study", components = 1:5))
dev.off()
```

## Normal tissue samples appear to be skewing the results, contributing a significant proporiton to overall variance. Try removing and rerunning analysis.

```{r}
# Subset metadata to remove normal tissue samples
metadata2 <- metadata %>% dplyr::filter(!Source == "normal")
# Subset counts matrix for samples to be assayed
matrix_ordered2 <- matrix_ordered[,c(rownames(metadata2))]
# Check counts matrix colnames match metadata rownames
all(rownames(metadata2) == colnames(matrix_ordered2))
```

```{r}
# Filter counts by cpm
cpms2 <- edgeR::cpm(matrix_ordered2)
keep2 <- rowSums(cpms2 >= 1 ) >= 2
cpms2 <- cpms2[keep2,]
```

## Principal Component Analysis of cpms

```{r}
p2 <- pca(cpms2, metadata = metadata2, removeVar = 0.1)
```

```{r}
# Screeplot
scree2.pdf <- paste0("Scree_cpm1_filtered_minus_normal.pdf")
pdf(file = file.path(resultsDir, scree2.pdf), width = 12, height = 8)
print(screeplot(p2, axisLabSize = 18, titleLabSize = 22, components = 1:10))
dev.off()

scree2.svg <- paste0("Scree_cpm1_filtered_minus_normal.svg")
pdf(file = file.path(resultsDir, scree2.svg), width = 12, height = 8)
print(screeplot(p2, axisLabSize = 18, titleLabSize = 22, components = 1:10))
dev.off()
```

```{r}
# PCA plot
pcabiplot2.pdf <- paste0("PCA_biplot_cpm1_filtered_minus_normal.pdf")
pdf(file = file.path(resultsDir, pcabiplot2.pdf), width = 12, height = 12)
print(biplot(p2, x = x, y = y, colby = "Source", shape = "Study", hline = 0, vline = 0,
             legendPosition = 'right', 
             xlab = paste0(x, ", ", round(p2$variance[x], digits = 2), "%"), 
             ylab = paste0(y, ", ", round(p2$variance[y], digits = 2), "%"), lab = NULL))
dev.off()

pcabiplot2.svg <- paste0("PCA_biplot_cpm1_filtered_minus_normal.svg")
svg(file = file.path(resultsDir, pcabiplot2.svg), width = 12, height = 12)
print(biplot(p2, x = x, y = y, colby = "Source", shape = "Study", hline = 0, vline = 0,
             legendPosition = 'right', 
             xlab = paste0(x, ", ", round(p2$variance[x], digits = 2), "%"), 
             ylab = paste0(y, ", ", round(p2$variance[y], digits = 2), "%"), lab = NULL))
dev.off()
```

```{r}
pairsplot2.pdf <- paste0("pairs_plot_PCA1-5_cpm1_filtered_minus_normal.pdf")
pdf(file = file.path(resultsDir, pairsplot2.pdf), width = 12, height = 12)
print(pairsplot(p2, colby = "Source", shape = "Study", components = 1:5))
dev.off()

pairsplot2.svg <- paste0("pairs_plot_PCA1-5_cpm1_filtered_minus_normal.svg")
pdf(file = file.path(resultsDir, pairsplot2.svg), width = 12, height = 12)
print(pairsplot(p2, colby = "Source", shape = "Study", components = 1:5))
dev.off()
```

```{r}
rm(p2)
```

## Principle component analysis of variance stabilising transformed data

```{r}
# Create deseq dataset
coldata2 <- metadata2
dds2 <- DESeqDataSetFromMatrix(countData = matrix_ordered2,
                              colData = coldata2,
                              design = ~ Metastatic)
# Set reference for comparison
dds2$Metastatic <- relevel(dds2$Metastatic, ref = "No")

# Prefilter low counts
dds2 <- estimateSizeFactors(dds2)
keep2 <- rowSums( counts(dds2, normalized=TRUE) >= 1 ) >= 2
dds2 <- dds2[keep2,]

# Run Deseq
dds2 <- DESeq(dds2)

# Run variance stabilising transform
vsd2 <- assay(vst(dds2))
```

```{r}
p2 <- pca(vsd2, metadata = colData(dds2), removeVar = 0.1)
```

```{r}
# Screeplot
scree2.pdf <- paste0("Scree_vst_minus_normal.pdf")
pdf(file = file.path(resultsDir, scree2.pdf), width = 12, height = 8)
print(screeplot(p2, axisLabSize = 18, titleLabSize = 22, components = 1:10))
dev.off()

scree2.svg <- paste0("Scree_vst_minus_normal.svg")
pdf(file = file.path(resultsDir, scree2.svg), width = 12, height = 8)
print(screeplot(p2, axisLabSize = 18, titleLabSize = 22, components = 1:10))
dev.off()
```

```{r}
# PCA plot
x <- paste0("PC1")
y <- paste0("PC2")

pcabiplot2.pdf <- paste0("PCA_biplot_vst_minus_normal_PC1_PC2.pdf")
pdf(file = file.path(resultsDir, pcabiplot2.pdf), width = 10, height = 6)
print(biplot(p2, x = x, y = y, colby = "Source", shape = "Study", hline = 0, vline = 0,
             legendPosition = 'right', 
             xlab = paste0(x, ", ", round(p2$variance[x], digits = 2), "%"), 
             ylab = paste0(y, ", ", round(p2$variance[y], digits = 2), "%"), lab = NULL))
dev.off()

pcabiplot2.svg <- paste0("PCA_biplot_vst_minus_normal_PC1_PC2.svg")
svg(file = file.path(resultsDir, pcabiplot2.svg), width = 10, height = 6)
print(biplot(p2, x = x, y = y, colby = "Source", shape = "Study", hline = 0, vline = 0,
             legendPosition = 'right', 
             xlab = paste0(x, ", ", round(p2$variance[x], digits = 2), "%"), 
             ylab = paste0(y, ", ", round(p2$variance[y], digits = 2), "%"), lab = NULL))
dev.off()



x <- paste0("PC2")
y <- paste0("PC3")

pcabiplot2.pdf <- paste0("PCA_biplot_vst_minus_normal_PC2_PC3.pdf")
pdf(file = file.path(resultsDir, pcabiplot2.pdf), width = 10, height = 6)
print(biplot(p2, x = x, y = y, colby = "Source", shape = "Study", hline = 0, vline = 0,
             legendPosition = 'right', 
             xlab = paste0(x, ", ", round(p2$variance[x], digits = 2), "%"), 
             ylab = paste0(y, ", ", round(p2$variance[y], digits = 2), "%"), lab = NULL))
dev.off()

pcabiplot2.svg <- paste0("PCA_biplot_vst_minus_normal_PC2_PC3.svg")
svg(file = file.path(resultsDir, pcabiplot2.svg), width = 10, height = 6)
print(biplot(p2, x = x, y = y, colby = "Source", shape = "Study", hline = 0, vline = 0,
             legendPosition = 'right', 
             xlab = paste0(x, ", ", round(p2$variance[x], digits = 2), "%"), 
             ylab = paste0(y, ", ", round(p2$variance[y], digits = 2), "%"), lab = NULL))
dev.off()



x <- paste0("PC3")
y <- paste0("PC4")

pcabiplot2.pdf <- paste0("PCA_biplot_vst_minus_normal_PC3_PC4.pdf")
pdf(file = file.path(resultsDir, pcabiplot2.pdf), width = 10, height = 6)
print(biplot(p2, x = x, y = y, colby = "Source", shape = "Study", hline = 0, vline = 0,
             legendPosition = 'right', 
             xlab = paste0(x, ", ", round(p2$variance[x], digits = 2), "%"), 
             ylab = paste0(y, ", ", round(p2$variance[y], digits = 2), "%"), lab = NULL))
dev.off()

pcabiplot2.svg <- paste0("PCA_biplot_vst_minus_normal_PC3_PC4.svg")
svg(file = file.path(resultsDir, pcabiplot2.svg), width = 10, height = 6)
print(biplot(p2, x = x, y = y, colby = "Source", shape = "Study", hline = 0, vline = 0,
             legendPosition = 'right', 
             xlab = paste0(x, ", ", round(p2$variance[x], digits = 2), "%"), 
             ylab = paste0(y, ", ", round(p2$variance[y], digits = 2), "%"), lab = NULL))
dev.off()
```

```{r}
pairsplot2.pdf <- paste0("pairs_plot_PCA1-5_vst_minus_normal.pdf")
pdf(file = file.path(resultsDir, pairsplot2.pdf), width = 10, height = 10)
print(pairsplot(p2, colby = "Source", shape = "Study", components = 1:5))
dev.off()

pairsplot2.svg <- paste0("pairs_plot_PCA1-5_vst_minus_normal.svg")
pdf(file = file.path(resultsDir, pairsplot2.svg), width = 10, height = 10)
print(pairsplot(p2, colby = "Source", shape = "Study", components = 1:5))
dev.off()
```

## Cell line only PCA

Rerun PCA using only 786 and OS cell lines (no primary tissue)

```{r}
# Subset metadata to remove normal tissue samples
metadata3 <- metadata %>% dplyr::filter(source_type == "Cell line")
# Subset counts matrix for samples to be assayed
matrix_ordered3 <- matrix_ordered[,c(rownames(metadata3))]
# Check counts matrix colnames match metadata rownames
all(rownames(metadata3) == colnames(matrix_ordered3))
```

```{r}
# Create deseq dataset
coldata3 <- metadata3
dds3 <- DESeqDataSetFromMatrix(countData = matrix_ordered3,
                              colData = coldata3,
                              design = ~ Metastatic)
# Set reference for comparison
dds3$Metastatic <- relevel(dds3$Metastatic, ref = "No")

# Prefilter low counts
dds3 <- estimateSizeFactors(dds3)
keep3 <- rowSums( counts(dds3, normalized=TRUE) >= 1 ) >= 2
dds3 <- dds3[keep3,]

# Run Deseq
dds3 <- DESeq(dds3)

# Run variance stabilising transform
vsd3 <- assay(vst(dds3))
```

```{r}
p2 <- pca(vsd3, metadata = colData(dds3), removeVar = 0.1)
```

```{r}
# Screeplot
scree3.pdf <- paste0("Scree_vst_cell_line.pdf")
pdf(file = file.path(resultsDir, scree3.pdf), width = 12, height = 8)
print(screeplot(p2, axisLabSize = 18, titleLabSize = 22, components = 1:10))
dev.off()

scree3.svg <- paste0("Scree_vst_cell_line.svg")
pdf(file = file.path(resultsDir, scree3.svg), width = 12, height = 8)
print(screeplot(p2, axisLabSize = 18, titleLabSize = 22, components = 1:10))
dev.off()
```

```{r}
# PCA plot
x <- paste0("PC1")
y <- paste0("PC2")

pcabiplot2.pdf <- paste0("PCA_biplot_vst_cell_line_PC1_PC2.pdf")
pdf(file = file.path(resultsDir, pcabiplot2.pdf), width = 10, height = 6)
print(biplot(p2, x = x, y = y, colby = "Source", shape = "Study", hline = 0, vline = 0,
             legendPosition = 'right', 
             xlab = paste0(x, ", ", round(p2$variance[x], digits = 2), "%"), 
             ylab = paste0(y, ", ", round(p2$variance[y], digits = 2), "%"), lab = NULL))
dev.off()

pcabiplot2.svg <- paste0("PCA_biplot_vst_cell_line_PC1_PC2.svg")
svg(file = file.path(resultsDir, pcabiplot2.svg), width = 10, height = 6)
print(biplot(p2, x = x, y = y, colby = "Source", shape = "Study", hline = 0, vline = 0,
             legendPosition = 'right', 
             xlab = paste0(x, ", ", round(p2$variance[x], digits = 2), "%"), 
             ylab = paste0(y, ", ", round(p2$variance[y], digits = 2), "%"), lab = NULL))
dev.off()



x <- paste0("PC2")
y <- paste0("PC3")

pcabiplot2.pdf <- paste0("PCA_biplot_vst_cell_line_PC2_PC3.pdf")
pdf(file = file.path(resultsDir, pcabiplot2.pdf), width = 10, height = 6)
print(biplot(p2, x = x, y = y, colby = "Source", shape = "Study", hline = 0, vline = 0,
             legendPosition = 'right', 
             xlab = paste0(x, ", ", round(p2$variance[x], digits = 2), "%"), 
             ylab = paste0(y, ", ", round(p2$variance[y], digits = 2), "%"), lab = NULL))
dev.off()

pcabiplot2.svg <- paste0("PCA_biplot_vst_cell_line_PC2_PC3.svg")
svg(file = file.path(resultsDir, pcabiplot2.svg), width = 10, height = 6)
print(biplot(p2, x = x, y = y, colby = "Source", shape = "Study", hline = 0, vline = 0,
             legendPosition = 'right', 
             xlab = paste0(x, ", ", round(p2$variance[x], digits = 2), "%"), 
             ylab = paste0(y, ", ", round(p2$variance[y], digits = 2), "%"), lab = NULL))
dev.off()



x <- paste0("PC3")
y <- paste0("PC4")

pcabiplot2.pdf <- paste0("PCA_biplot_vst_cell_line_PC3_PC4.pdf")
pdf(file = file.path(resultsDir, pcabiplot2.pdf), width = 10, height = 6)
print(biplot(p2, x = x, y = y, colby = "Source", shape = "Study", hline = 0, vline = 0,
             legendPosition = 'right', 
             xlab = paste0(x, ", ", round(p2$variance[x], digits = 2), "%"), 
             ylab = paste0(y, ", ", round(p2$variance[y], digits = 2), "%"), lab = NULL))
dev.off()

pcabiplot2.svg <- paste0("PCA_biplot_vst_cell_line_PC3_PC4.svg")
svg(file = file.path(resultsDir, pcabiplot2.svg), width = 10, height = 6)
print(biplot(p2, x = x, y = y, colby = "Source", shape = "Study", hline = 0, vline = 0,
             legendPosition = 'right', 
             xlab = paste0(x, ", ", round(p2$variance[x], digits = 2), "%"), 
             ylab = paste0(y, ", ", round(p2$variance[y], digits = 2), "%"), lab = NULL))
dev.off()
```

```{r}
pairsplot2.pdf <- paste0("pairs_plot_PCA1-5_vst_cell_line.pdf")
pdf(file = file.path(resultsDir, pairsplot2.pdf), width = 10, height = 10)
print(pairsplot(p2, colby = "Source", shape = "Study", components = 1:5))
dev.off()

pairsplot2.svg <- paste0("pairs_plot_PCA1-5_vst_cell_line.svg")
pdf(file = file.path(resultsDir, pairsplot2.svg), width = 10, height = 10)
print(pairsplot(p2, colby = "Source", shape = "Study", components = 1:5))
dev.off()
```

# Rodrigues et al cell lines only

```{r}
# Subset metadata to remove normal tissue samples
metadata4 <- metadata %>% dplyr::filter(Study == "Rodrigues, P., et al (2018). Cancer discovery")
# Subset counts matrix for samples to be assayed
matrix_ordered4 <- matrix_ordered[,c(rownames(metadata4))]
# Check counts matrix colnames match metadata rownames
all(rownames(metadata4) == colnames(matrix_ordered4))
```

```{r}
# Create deseq dataset
coldata4 <- metadata4
dds4 <- DESeqDataSetFromMatrix(countData = matrix_ordered4,
                              colData = coldata4,
                              design = ~ Metastatic)
# Set reference for comparison
dds4$Metastatic <- relevel(dds4$Metastatic, ref = "No")

# Prefilter low counts
dds4 <- estimateSizeFactors(dds4)
keep4 <- rowSums( counts(dds4, normalized=TRUE) >= 1 ) >= 2
dds4 <- dds4[keep4,]

# Run Deseq
dds4 <- DESeq(dds4)

# Run variance stabilising transform
vsd4 <- assay(vst(dds4))
```

```{r}
p2 <- pca(vsd4, metadata = colData(dds4), removeVar = 0.1)
```

```{r}
# Screeplot
scree3.pdf <- paste0("Scree_vst_Rodrigues.pdf")
pdf(file = file.path(resultsDir, scree3.pdf), width = 12, height = 8)
print(screeplot(p2, axisLabSize = 18, titleLabSize = 22, components = 1:10))
dev.off()

scree3.svg <- paste0("Scree_vst_Rodrigues.svg")
pdf(file = file.path(resultsDir, scree3.svg), width = 12, height = 8)
print(screeplot(p2, axisLabSize = 18, titleLabSize = 22, components = 1:10))
dev.off()
```

```{r}
# PCA plot
x <- paste0("PC1")
y <- paste0("PC2")

pcabiplot2.pdf <- paste0("PCA_biplot_vst_Rodrigues_PC1_PC2.pdf")
pdf(file = file.path(resultsDir, pcabiplot2.pdf), width = 7, height = 6)
print(biplot(p2, x = x, y = y, colby = "Source", hline = 0, vline = 0,
             legendPosition = 'right', 
             xlab = paste0(x, ", ", round(p2$variance[x], digits = 2), "%"), 
             ylab = paste0(y, ", ", round(p2$variance[y], digits = 2), "%"), lab = NULL))
dev.off()

pcabiplot2.svg <- paste0("PCA_biplot_vst_Rodrigues_PC1_PC2.svg")
svg(file = file.path(resultsDir, pcabiplot2.svg), width = 7, height = 6)
print(biplot(p2, x = x, y = y, colby = "Source", hline = 0, vline = 0,
             legendPosition = 'right', 
             xlab = paste0(x, ", ", round(p2$variance[x], digits = 2), "%"), 
             ylab = paste0(y, ", ", round(p2$variance[y], digits = 2), "%"), lab = NULL))
dev.off()



x <- paste0("PC2")
y <- paste0("PC3")

pcabiplot2.pdf <- paste0("PCA_biplot_vst_Rodrigues_PC2_PC3.pdf")
pdf(file = file.path(resultsDir, pcabiplot2.pdf), width = 7, height = 6)
print(biplot(p2, x = x, y = y, colby = "Source", hline = 0, vline = 0,
             legendPosition = 'right', 
             xlab = paste0(x, ", ", round(p2$variance[x], digits = 2), "%"), 
             ylab = paste0(y, ", ", round(p2$variance[y], digits = 2), "%"), lab = NULL))
dev.off()

pcabiplot2.svg <- paste0("PCA_biplot_vst_Rodrigues_PC2_PC3.svg")
svg(file = file.path(resultsDir, pcabiplot2.svg), width = 7, height = 6)
print(biplot(p2, x = x, y = y, colby = "Source", hline = 0, vline = 0,
             legendPosition = 'right', 
             xlab = paste0(x, ", ", round(p2$variance[x], digits = 2), "%"), 
             ylab = paste0(y, ", ", round(p2$variance[y], digits = 2), "%"), lab = NULL))
dev.off()



x <- paste0("PC3")
y <- paste0("PC4")

pcabiplot2.pdf <- paste0("PCA_biplot_vst_Rodrigues_PC3_PC4.pdf")
pdf(file = file.path(resultsDir, pcabiplot2.pdf), width = 7, height = 6)
print(biplot(p2, x = x, y = y, colby = "Source", hline = 0, vline = 0,
             legendPosition = 'right', 
             xlab = paste0(x, ", ", round(p2$variance[x], digits = 2), "%"), 
             ylab = paste0(y, ", ", round(p2$variance[y], digits = 2), "%"), lab = NULL))
dev.off()

pcabiplot2.svg <- paste0("PCA_biplot_vst_Rodrigues_PC3_PC4.svg")
svg(file = file.path(resultsDir, pcabiplot2.svg), width = 7, height = 6)
print(biplot(p2, x = x, y = y, colby = "Source", hline = 0, vline = 0,
             legendPosition = 'right', 
             xlab = paste0(x, ", ", round(p2$variance[x], digits = 2), "%"), 
             ylab = paste0(y, ", ", round(p2$variance[y], digits = 2), "%"), lab = NULL))
dev.off()
```

```{r}
pairsplot2.pdf <- paste0("pairs_plot_PCA1-5_vst_Rodrigues.pdf")
pdf(file = file.path(resultsDir, pairsplot2.pdf), width = 10, height = 10)
print(pairsplot(p2, colby = "Source", components = 1:5))
dev.off()

pairsplot2.svg <- paste0("pairs_plot_PCA1-5_vst_Rodrigues.svg")
pdf(file = file.path(resultsDir, pairsplot2.svg), width = 10, height = 10)
print(pairsplot(p2, colby = "Source", components = 1:5))
dev.off()
```


# Rodrigues vs Liu stage 1,3,4

```{r}
# Subset metadata to remove normal tissue samples
metadata4 <- metadata %>% dplyr::filter(Study == "Rodrigues, P., et al (2018). Cancer discovery" | Study == "Le, L. N., et al. (2024). Frontiers in oncology")
metadata4 <- metadata4 %>% dplyr::filter(!Source == "normal", !tumor_stage == "2", ! tumor_stage == "3")
# Subset counts matrix for samples to be assayed
matrix_ordered4 <- matrix_ordered[,c(rownames(metadata4))]
# Check counts matrix colnames match metadata rownames
all(rownames(metadata4) == colnames(matrix_ordered4))
```

```{r}
# Create deseq dataset
coldata4 <- metadata4
dds4 <- DESeqDataSetFromMatrix(countData = matrix_ordered4,
                              colData = coldata4,
                              design = ~ Metastatic)
# Set reference for comparison
dds4$Metastatic <- relevel(dds4$Metastatic, ref = "No")

# Prefilter low counts
dds4 <- estimateSizeFactors(dds4)
keep4 <- rowSums( counts(dds4, normalized=TRUE) >= 1 ) >= 2
dds4 <- dds4[keep4,]

# Run Deseq
dds4 <- DESeq(dds4)

# Run variance stabilising transform
vsd4 <- assay(vst(dds4))
```

```{r}
p2 <- pca(vsd4, metadata = colData(dds4), removeVar = 0.1)
```

```{r}
# Screeplot
scree3.pdf <- paste0("Scree_vst_Rodrigues_vs_Liu.pdf")
pdf(file = file.path(resultsDir, scree3.pdf), width = 12, height = 8)
print(screeplot(p2, axisLabSize = 18, titleLabSize = 22, components = 1:10))
dev.off()

scree3.svg <- paste0("Scree_vst_Rodrigues_vs_Liu.svg")
pdf(file = file.path(resultsDir, scree3.svg), width = 12, height = 8)
print(screeplot(p2, axisLabSize = 18, titleLabSize = 22, components = 1:10))
dev.off()
```

```{r}
# PCA plot
x <- paste0("PC1")
y <- paste0("PC2")

pcabiplot2.pdf <- paste0("PCA_biplot_vst_Rodrigues_vs_Liu_PC1_PC2.pdf")
pdf(file = file.path(resultsDir, pcabiplot2.pdf), width = 7, height = 6)
print(biplot(p2, x = x, y = y, colby = "Source", shape = "source_type", hline = 0, vline = 0,
             legendPosition = 'right', 
             xlab = paste0(x, ", ", round(p2$variance[x], digits = 2), "%"), 
             ylab = paste0(y, ", ", round(p2$variance[y], digits = 2), "%"), lab = NULL))
dev.off()

pcabiplot2.svg <- paste0("PCA_biplot_vst_Rodrigues_vs_Liu_PC1_PC2.svg")
svg(file = file.path(resultsDir, pcabiplot2.svg), width = 7, height = 6)
print(biplot(p2, x = x, y = y, colby = "Source", shape = "source_type",hline = 0, vline = 0,
             legendPosition = 'right', 
             xlab = paste0(x, ", ", round(p2$variance[x], digits = 2), "%"), 
             ylab = paste0(y, ", ", round(p2$variance[y], digits = 2), "%"), lab = NULL))
dev.off()



x <- paste0("PC2")
y <- paste0("PC3")

pcabiplot2.pdf <- paste0("PCA_biplot_vst_Rodrigues_vs_Liu_PC2_PC3.pdf")
pdf(file = file.path(resultsDir, pcabiplot2.pdf), width = 7, height = 6)
print(biplot(p2, x = x, y = y, colby = "Source", shape = "source_type",hline = 0, vline = 0,
             legendPosition = 'right', 
             xlab = paste0(x, ", ", round(p2$variance[x], digits = 2), "%"), 
             ylab = paste0(y, ", ", round(p2$variance[y], digits = 2), "%"), lab = NULL))
dev.off()

pcabiplot2.svg <- paste0("PCA_biplot_vst_Rodrigues_vs_Liu_PC2_PC3.svg")
svg(file = file.path(resultsDir, pcabiplot2.svg), width = 7, height = 6)
print(biplot(p2, x = x, y = y, colby = "Source", shape = "source_type",hline = 0, vline = 0,
             legendPosition = 'right', 
             xlab = paste0(x, ", ", round(p2$variance[x], digits = 2), "%"), 
             ylab = paste0(y, ", ", round(p2$variance[y], digits = 2), "%"), lab = NULL))
dev.off()



x <- paste0("PC3")
y <- paste0("PC4")

pcabiplot2.pdf <- paste0("PCA_biplot_vst_Rodrigues_vs_Liu_PC3_PC4.pdf")
pdf(file = file.path(resultsDir, pcabiplot2.pdf), width = 7, height = 6)
print(biplot(p2, x = x, y = y, colby = "Source", shape = "source_type",hline = 0, vline = 0,
             legendPosition = 'right', 
             xlab = paste0(x, ", ", round(p2$variance[x], digits = 2), "%"), 
             ylab = paste0(y, ", ", round(p2$variance[y], digits = 2), "%"), lab = NULL))
dev.off()

pcabiplot2.svg <- paste0("PCA_biplot_vst_Rodrigues_vs_Liu_PC3_PC4.svg")
svg(file = file.path(resultsDir, pcabiplot2.svg), width = 7, height = 6)
print(biplot(p2, x = x, y = y, colby = "Source", shape = "source_type",hline = 0, vline = 0,
             legendPosition = 'right', 
             xlab = paste0(x, ", ", round(p2$variance[x], digits = 2), "%"), 
             ylab = paste0(y, ", ", round(p2$variance[y], digits = 2), "%"), lab = NULL))
dev.off()
```

```{r}
pairsplot2.pdf <- paste0("pairs_plot_PCA1-5_vst_Rodrigues_vs_Liu.pdf")
pdf(file = file.path(resultsDir, pairsplot2.pdf), width = 10, height = 10)
print(pairsplot(p2, colby = "Source", components = 1:5))
dev.off()

pairsplot2.svg <- paste0("pairs_plot_PCA1-5_vst_Rodrigues_vs_Liu.svg")
pdf(file = file.path(resultsDir, pairsplot2.svg), width = 10, height = 10)
print(pairsplot(p2, colby = "Source", components = 1:5))
dev.off()
```